---
title: "Sick Cats Project"
author: "Ana Sofia Guerra & Gabriel Runte"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello! Testing this. Try Comitting and pushing with a response to make sure we get the hang of this. 
 
```{r packages}
#install.packages("deSolve", dependencies = T)
library(deSolve)
require(deSolve)

library(tidyverse) #install.packages('tidyverse)
```
 
# Written Paper

### Introduction
200 words, at least two references. 
 
## The Model

**1. Describe core model equations in words**

starts with Lotka-Volterra predation model:
$$
\begin{align}
\frac{d N}{dT}  &= rN(1-N/K)-\mu NP 
\newline
\newline
\frac{d P}{dT} &= a\mu NP-dP
\newline
\end{align}
$$
where:
N(T): population size of prey (birds)
P(T): population size of predators (cats)
T: time
*r*: per capita growth rate
*K*: carrying capacity
*mu*: predation rate (per capita and unit prey)
*d*: predator per-capita mortality rate
*a*: trophic conversion efficiency

...incorporating FIV, P = Susceptible + Infected (with no recovery or immunity), and there is only one stage of the illness (FIV). so:

$$
\begin{align}
\ P= S + I
\newline
\newline
\frac{d S}{dT}  &=  a\mu NP-dS-\theta(P) \frac{S I}{P}
\newline
\newline
\frac{d I}{dT} &= \theta(P) \frac{S I}{P}-dI-\delta I
\newline
\end{align}
$$
where: $$\theta(P) = \beta_{PM}$$ is the incidence is of proportionate mixing type if the contact rate between predator individuals in constant.

The incidence is of mass actioin type if the conact rate between predator individuals increases linearly with population size, $$\theta(P) = \beta_{PM} P$$


re-writing model equations, with i being disease prevalence: 
$$ i = I/(S+I) $$

They introduce dimensionless variables (in order to reduce the number of parameters), where:
$$ 
\begin{align}
B=N/K
\newline
C=P\mu /r
\newline
t=rT
\end{align}
$$
**so the main equations in the model are: **

$$
\begin{align}
\frac{d B}{dt}  &= B(1-B)-BC
\newline
\newline
\frac{d C}{dt}  &= eBC-mC-\alpha Ci
\newline
\newline
\frac{d i}{dt}  &= ([\Gamma(C) - \alpha] [1-i] - eB)i
\newline
\end{align}
$$
where

$$
\begin{align}
\Gamma (C)  &= \frac{\beta_{PM}}{r} = \sigma_{PM}
\newline
\newline
and
\newline
\newline
\Gamma (C)  &= \frac{\beta_{MA}}{\mu}C = \sigma_{MA}C
\newline
\end{align}
$$
for proportionate mixing and mass action transmission respectively. 

$$
\begin{align}
e  &= \frac{\alpha \mu K}{r}
\newline
\newline
m  &= \frac{d}{r}
\newline
\newline
\alpha  &= \frac{\delta}{r}
\newline
\end{align}
$$

#### Coding the model

```{r known.parameters}
#known model parameters
r <- 0.1 #bird growth rate
a <- 0.03 #trophic conversion efficiency
d <-  0.6 #(cat natural death rate) 
BPM <- 1.6 #constant contact rate
v <-  0.2 #virulence
```

```{r unknown.parameters}
K <- #some value here #bird carrying capacity
mu <- #some value #cat predation rate
BMA <- #some value #contact rate2
```

```{r dimensionless.parameters}
e <- (a*mu*K)/r
m <- d/r
al <- v/r
```


```{r variables}
tset <- seq(from = 0, to = 60, length.out = 10000)
B.sim <- NaN*tset #bird pop 
C.sim <- NaN*tset #cat pop 
i.sim <- NaN*tset #disease prevalence
```





### Model Assumptions:
List model assumptions

### Variables and Parameters
Table of variables and parameters, meanings, units, and any values used in paper

```{r parameters.table, include=T}
#parameters.table <- data.frame("parameter" = c("r","K",paste($\mu$),"a","d",), "definition" = " ", "value" = , "units" =, stringsAsFactors = FALSE)
```

### Time Series
Show how outcomes differ as a function of interesting parameters. include caption. 

### Bifurcation diagram

## Extension Section

1. Describe a modification to the model - in words and in math

2. Illustrate effects of modiification with one figure 

### Discussion

1. 100 words + 2 references

2. Critique applicability/limitations

3. Interpret results - what is the ecological significance? 


