---
title: "Sick Cats & Mathematical Modeling"
subtitle: "Final Project for EEMB279"
author: "Ana Sofia Guerra & Gabriel Runte"
output: word_document
---

 
```{r packages, include=FALSE, echo=FALSE}
#install.packages("deSolve", dependencies = T)
library(deSolve)
require(deSolve)
library(tidyverse) #install.packages('tidyverse)
library(kableExtra) #install.packages('kableExtra')
```
 



# A Chalk Talk Abstract

<<Insert our own here.>>

# Introduction
<<Insert Introduction from Google Doc>>
  
# The Model
  
A story of sick cats based on the mathematical model by: Oliveira NM & Hilker FM (2010), Modelling disease introduction as biological control of invasive predators to preserve endangered prey, **Bulletin of Mathematical Biology**, 72:444-468. 
  
Oliveira and Hilker base their model on the Lotka-Volterra predation model, where a logistically-growing bird population (N) is predated on by introduced cats (P).  
$$
\begin{align}
\frac{d N}{dT}  &= rN(1-N/K)-\mu NP 
\newline
\newline
\frac{d P}{dT} &= a\mu NP-dP
\newline
\end{align}
$$
Here,  
N(T): population size of prey (birds)
P(T): population size of predators (cats)
T: time
*r*: per capita growth rate
*K*: carrying capacity
*mu*: predation rate (per capita and unit prey)
*d*: predator per-capita mortality rate
*a*: trophic conversion efficiency
  
The introduction of FIV into the cat population is modeled where P = Susceptible cats + Infected cats (with no recovery or immunity), and there is only one stage of the illness (FIV).  
  
$$
\begin{align}
\ P= S + I
\newline
\newline
\frac{d S}{dT}  &=  a\mu NP-dS-\theta(P) \frac{S I}{P}
\newline
\newline
\frac{d I}{dT} &= \theta(P) \frac{S I}{P}-dI-\delta I
\newline
\end{align}
$$
  
where: $$\theta(P) = \beta_{PM}$$ is the incidence is of proportionate mixing type if the contact rate between predator individuals in constant, which is what we will focus on for the purposes of this assignmant. 
The disease equations can be rearranged to represent disease prevalence (i):  
$$ i = I/(S+I) $$  
  
As many of the parameter values for the original equations are unknown, Oliveira and Hilker inistead opted for  introducing dimensionless variables (in order to reduce the number of parameters), where:  
$$ 
\begin{align}
B=N/K
\newline
C=P\mu /r
\newline
t=rT
\end{align}
$$
  
The models then become:  
  
$$
\begin{align}
\frac{d B}{dt}  &= B(1-B)-BC
\newline
\newline
\frac{d C}{dt}  &= eBC-mC-\alpha Ci
\newline
\newline
\frac{d i}{dt}  &= ([\Gamma(C) - \alpha] [1-i] - eB)i
\newline
\end{align}
$$
where  
  
$$
\begin{align}
\Gamma (C)  &= \frac{\beta_{PM}}{r} = \sigma_{PM}
\newline
\newline
\end{align}
$$
for a proportionate mixing scenario. These models can be roughly interpreted as $$dB/dt$$ being the number of birds relative to their carrying capacity, $$dC/dt$$ is the cat population relative to their attack rate on biirds and the growth rate of birds, and $$di/dt$$ is the prevalence of FIV.  
The parameters in the nondimensional equations are calculated as:  
$$
\begin{align}
e  &= \frac{a \mu K}{r}
\newline
\newline
m  &= \frac{d}{r}
\newline
\newline
\alpha  &= \frac{\delta}{r}
\newline
\end{align}
$$
where $$e$$ can be considered cats' conversion efficiency/attack rate, $$m$$ is natural cat death, and $$\alpha$$ is sick cat death.  
  
## Model Assumptions:
<<Insert Assumptions from Google Doc >>

## Variables and Parameters
  
### Dimensional Equations
Below are the values and descriptions for the vairables and parameters found in the original dimensional equations:  

```{r d.variables.table, include=T}
dv <- data.frame("variable" = c("N","P","S","I"), "definition" = c("bird population","total cat population","susceptible cat","infected cats"), stringsAsFactors = FALSE)

kable(dv) %>% kable_styling() 
```

```{r d.parameters.table, include=T}
##have to figure out how to get the greek letter parameters to show. 
dp <- data.frame("parameter" = c("r","K", "$\\\\mu$","a","d", "$\\\\beta$"
, "$\\\\delta$"), "definition" = c("per capita growth rate of birds","carrying capacity of birds","predation rate","trophic conversion efficiency","natural per-capita death rate of cats","proportionate mixing contact rate","virulence"), "value" = c("0.1","-","-","0.03","0.6","1.5","0.2"), stringsAsFactors = FALSE)

kable(dp) %>% kable_styling() 
```
  
However, the authors opted against using the dimensional equations as many of the parameters are unknown. Therefore, they used the variables and parameters below, and chose other biologically relevant values for one of the more unknown parameters (e.g. **e**).  
```{r d.variables.table, include=T}
ndv <- data.frame("variable" = c("B","C","i"), "definition" = c("bird numbers relative to K","Cat numbers relative to attack rate and bird population growth","FIV prevalence"), stringsAsFactors = FALSE)

kable(ndv) %>% kable_styling() 
```  
```{r d.parameters.table, include=T}
##have to figure out how to get the greek letter parameters to show. 
ndp <- data.frame("parameter" = c("e","m", "$\\\\alpha$", "$\\\\sigma$"), "definition" = c("cat conversion efficiency/attack rate","natura cat death","infected cat death","contact rate"), "value" = c("10","6","2","15"), stringsAsFactors = FALSE)

kable(ndp) %>% kable_styling() 
```


### Time Series
Show how outcomes differ as a function of interesting parameters. include caption. 

### Bifurcation diagram

## Extension Section

1. Describe a modification to the model - in words and in math

2. Illustrate effects of modiification with one figure 

### Discussion

1. 100 words + 2 references

2. Critique applicability/limitations

3. Interpret results - what is the ecological significance? 


