---
title: "EEMB 279 Project"
author: "Gabriel Runte and Ana Sofia Guerra"
date: "2/18/2020"
output: html_document
subtitle: Time Series
---

A time series simulation of the model by Oliveira NM & Hilker FM (2010), Modelling disease introduction as biological control of invasive predators to preserve endangered prey, **Bulletin of Mathematical Biology**, 72:444-468. 

In the case of this model, the authors introduced dimensionless parameters to simply the model and thus avoid making incorrect assumptions about unknown variables such as$$\mu$$, the predation rate, and **K**, the bird population carrying capacity, as well as initial population sizes for birds, cats, and FIV. 

Thus, a lot of our values below are classified as `NaN` as they are not applied in the model, but do exist. 
```{r var.params, include=TRUE}
##Variables
#N.v <- NaN    ## population size for victims (birds)
#N.p <- NaN    ## population size for predators (cats)
t.time.set <- seq(1, 200, length.out= 2000) ## time

##Parameter values##
#K <- NaN            # carrying capacity               (will be estimated from dimensionless parameters)
#u <- NaN            # predation rate                  (will be estimated from dimensionless parameters)
#beta.ma <- NaN      # contact rate for mass action    (will be estimated from dimensionless parameters)    

r.v <- 0.1          # per capita growth rate of birds in the model
a <- 0.03           # trophic conversion efficiency of cats
d.p <- 0.6          # natural per capita death rate of cats
beta.pm <- 1.5      # contact rate for proportionate mixing
delta <- 0.2        # virulence

#core equations
#B <- N.v/K
B.set <-seq(0, 1, length.out=20000)
#C <- N.p*u/r.v
t.set <- r.v*t.time.set

#dimensionless parameters
sigma.pm <- beta.pm/r.v
#sigma.pm.set <- seq(0,25, length.out = 20000)
#sigma.ma <- beta.ma * C/u 
#sigma.ma.set<- seq(0, 250, length.out = 20000)

#e <- alpha*u*K/r.v
e <- 10
#e.set <- seq(0, 25, length.out = 20000)
m <- d.p/r.v
alpha <- delta/r.v

```

In this time series we selected a biologically relevant, yet random value for e (10 in this case), from the range provided by the authors of the study. We can produce a time series with different values of e. In this simulation, the disease is present at the beginning. 
```{r Simu1, include=T}
B.simu1 <- NaN*t.set 
  B.simu1[1] <- .3
C.simu1 <- NaN*t.set
  C.simu1[1] <- .75
I.simu1 <- NaN*t.set
 I.simu1[1] <- 0.2

#for PM
for(i in 2:length(t.set)){
  
  dt <- t.set[i]-t.set[i-1]
  B <- B.simu1[i-1]
  C <- C.simu1[i-1]
  I <- I.simu1[i-1]
  
  dB <- (B*(1-B) -B*C) *dt
  dC <- (e*B*C - m*C - alpha*C*I)*dt
  dI <- (((sigma.pm - alpha)* (1-I) - e*B)*I) *dt
  
  B.simu1[i] <- B + dB
  C.simu1[i] <- C + dC
  I.simu1[i] <- I + dI
}

  
plot(t.set, B.simu1, type= "l", ylim= c(0, 1), ylab="Population (dimensionless)", xlab="time (rT)")  
lines(t.set,C.simu1,lwd=2,col='dodgerblue')
lines(t.set, I.simu1, lwd = 2, col= "coral3")
legend(x=16,y=1, col = c("black","dodgerblue","coral3"), lty=1,lwd=2,legend = c("birds", "cats", "FIV"))

```

As was done in a time series in the study, there was a lag in the introduction of FIV to the cat population. In this case we introduced FIV at time step 3. 

```{r simu2, include=TRUE}
B.simu2 <- NaN*t.set 
  B.simu2[1] <- .3
C.simu2 <- NaN*t.set
  C.simu2[1] <- .75
I.simu2 <- NaN*t.set
# I.simu2[1] <- 0
```

```{r simu2.loop, include=TRUE}
#for PM
for(i in 2:length(t.set)){
  
  	if(t.set[i] < 3){I.simu2[i-1] <- 0 } else { 
  	  if(t.set[i] == 3) {I.simu2[i-1] <- 0.2} else { I<- I.simu2[i-1]}}
	
  dt <- t.set[i]-t.set[i-1]
  B <- B.simu2[i-1]
  C <- C.simu2[i-1]
  #I.simu2[i-1]
  
  dB <- (B*(1-B) -B*C) *dt
  dC <- (e*B*C - m*C - alpha*C*I)*dt
  dI <- (((sigma.pm - alpha)* (1-I) - e*B)*I) *dt
  
  B.simu2[i] <- B + dB
  C.simu2[i] <- C + dC
  I.simu2[i] <- I + dI
}

plot(t.set, B.simu2, type= "l", ylim= c(0, 1), ylab="Population (dimensionless)", xlab="time (rT)")  
lines(t.set,C.simu2,lwd=2,col='dodgerblue')
lines(t.set, I.simu2, lwd = 2, col= "coral3")
legend(x=16,y=1, col = c("black","dodgerblue","coral3"), lty=1,lwd=2,legend = c("birds", "cats", "FIV"))
text(x=3.5, y=0.04, adj=0, label="FIV introduced")
```


```{r, trying to add decreased efficiency of hunting with FIV}
l <- 0.25 #percent activity of sick cats (lathargy)
#for PM
B.simu3 <- NaN*t.set 
  B.simu3[1] <- .3
C.simu3 <- NaN*t.set
  C.simu3[1] <- .75
I.simu3 <- NaN*t.set
#I.simu3[1] <- 0

for(i in 2:length(t.set)){
  
  	if(t.set[i] < 3){I.simu3[i-1] <- 0 } else { 
  	  if(t.set[i] == 3) {I.simu3[i-1] <- 0.2} else { I<- I.simu3[i-1]}}
	
  dt <- t.set[i]-t.set[i-1]
  B <- B.simu3[i-1]
  C <- C.simu3[i-1]
  #I.simu3[i-1]
  
  dB <- (B*(1-B) -B*C*(1-I) - l*B*I*C ) *dt
  dC <- (e*(B*C*(1-I) + l*B*I*C) - m*C - alpha*C*I)*dt
  dI <- (((sigma.pm - alpha)* (1-I) - e*B)*I) *dt
  
  B.simu3[i] <- B + dB
  C.simu3[i] <- C + dC
  I.simu3[i] <- I + dI
}

plot(t.set, B.simu3, type= "l", ylim= c(0, 1), xlim=c(0, 25), ylab="Population (dimensionless)", xlab="time (rT)", col='paleturquoise4')  
lines(t.set,C.simu3,lwd=2,col='gray21')
lines(t.set, I.simu3, lwd = 2, col= "sienna3")
lines(t.set,C.simu2,lwd=2, lty= 2, col='gray67')
lines(t.set, I.simu2, lwd = 2,lty= 2, col= "sienna1")
lines(t.set,B.simu2,lwd=2,lty= 2, col='paleturquoise3')

legend(x=17,y=1, col = c("paleturquoise4","paleturquoise3","sienna3", "sienna1", "gray21", "gray67"), lty=1,lwd=2,legend = c("birds w/ FIV effect", "regular birds", "FIV w/ FIV effect", " regular FIV", "cats w/ FIV effect", "regular cats"))
text(x=3.5, y=0.04, adj=0, label="FIV introduced")




plot(t.set, B.simu3,type= "l", ylim= c(0, 1), ylab="Population (dimensionless)", xlab="time (rT)", col= 'Dodgerblue2')
lines(t.set, B.simu2, lwd=2, col= 'coral3')



plot(t.set, B.simu3, type= "l", ylim= c(0, 1), lwd=2, ylab="Population (dimensionless)", xlab="time (rT)", col='paleturquoise4')  
lines(t.set,C.simu3,lwd=2,col='gray21')
lines(t.set, I.simu3, lwd = 2, col= "sienna3")

legend(x=16,y=1, col = c("paleturquoise4","gray21","sienna3"), lty=1,lwd=2,legend = c("birds", "cats", "FIV"))
text(x=3.5, y=0.04, adj=0, label="FIV introduced")

```

```{r}

mu <- 4 #not given
a <- 0.3 #trophic conversion
d <- 0.6 #per capita death
beta.pm <- 13 #pm contact rate
gamma <- 4.8 # virulence
K<- 1
r<- 0.6

l <- 0.25 #percent activity of sick cats (lathargy)
#for PM
N.simu4 <- NaN*t.time.set 
  N.simu4[1] <- .3
S.simu4 <- NaN*t.time.set
  S.simu4[1] <- .75
I.simu4 <- NaN*t.time.set
  I.simu4[1] <- .002

for(i in 2:length(t.time.set)){

	
  dt <- t.time.set[i]-t.time.set[i-1]
  N <- N.simu4[i-1]
  S <- S.simu4[i-1]
 I <- I.simu4[i-1]
  P <- S + I
  
  dN <- (r*N*(1-N/K)-mu*N*P) * dt
  dS <- ((a*mu*N*P)-(d*S) - (beta.pm*(S*I/P))) * dt
  dI <- (beta.pm* (S*I/P) - d*I -gamma*I) * dt
  
  N.simu4[i] <- N + dN
  S.simu4[i] <- S + dS
  I.simu4[i] <- I + dI
 }

plot(t.time.set, N.simu4, type= "l", ylim= c(0, 1), xlim=c(0, 100), ylab="Population (dimensionless)", xlab="time (rT)", col='paleturquoise4')  
lines(t.time.set,S.simu4,lwd=2,col='gray21')
lines(t.time.set, I.simu4, lwd = 2, col= "sienna3")


legend(x=17,y=1, col = c("paleturquoise4","sienna3", "gray21" ), lty=1,lwd=2,legend = c( "regular birds",  " regular FIV", "regular cats"))
text(x=3.5, y=0.04, adj=0, label="FIV introduced")





```

```{r}
#trying to do the delayed introduction of FIV


mu <- 4 #not given
a <- 0.3 #trophic conversion
d <- 0.6 #per capita death
beta.pm <- 13 #pm contact rate
gamma <- 4.8 # virulence
K<- 1
r<- 0.6

l <- 0.8 #percent activity of sick cats (lathargy)
#for PM
N.simu5 <- NaN*t.time.set 
  N.simu5[1] <- .6
S.simu5 <- NaN*t.time.set
  S.simu5[1] <- .3
I.simu5 <- NaN*t.time.set
# I.simu5[1] <- .002

for(i in 2:length(t.time.set)){
  
 if(t.time.set[i] < 20){I.simu5[i-1] <- 0 } else { 
  if(t.time.set[i] == 20) {I.simu5[i-1] <- 0.1} else { I<- I.simu5[i-1]}}
	
  dt <- t.time.set[i]-t.time.set[i-1]
  N <- N.simu5[i-1]
  S <- S.simu5[i-1]
 # I <- I.simu5[i-1]
  P <- S + I
  
  dN <- (r*N*(1-N/K)-mu*N*P) * dt
  dS <- ((a*mu*N*P)-(d*S) - (beta.pm*(S*I/P))) * dt
  dI <- (beta.pm* (S*I/P) - d*I -gamma*I) * dt
  
  N.simu5[i] <- N + dN
  S.simu5[i] <- S + dS
  I.simu5[i] <- I + dI
 }

plot(t.time.set, N.simu5, type= "l", ylim= c(0, 1), xlim=c(0, 100), ylab="Population (dimensionless)", xlab="time (rT)", col='paleturquoise4')  
lines(t.time.set,S.simu5,lwd=2,col='gray21')
lines(t.time.set, I.simu5, lwd = 2, col= "sienna3")


legend(x=17,y=1, col = c("paleturquoise4","sienna3", "gray21" ), lty=1,lwd=2,legend = c( "regular birds",  " regular FIV", "regular cats"))
text(x=3.5, y=0.04, adj=0, label="FIV introduced")





```

